---
title: "Summary Statistics"
subtitle: "STAT 240 - Fall 2025"
author: "Robert Sholl"
footer:  "[github.io/stat240_f25](https://rmshksu.github.io/stat240_f25/)"
logo: "images/logo.jpg"
# embed-resources: true
format: 
  revealjs: 
    theme: slides.scss
    transition: fade
    slide-number: true
editor: visual
execute:
  freeze: auto
---

# Math Review

## Rules

```{=html}
<script type="text/x-mathjax-config">
MathJax.Hub.Register.StartupHook("TeX Jax Ready",function () {
  MathJax.Hub.Insert(MathJax.InputJax.TeX.Definitions.macros,{
    cancel: ["Extension","cancel"],
    bcancel: ["Extension","cancel"],
    xcancel: ["Extension","cancel"],
    cancelto: ["Extension","cancel"]
  });
});
</script>
```

```{r message=FALSE, warning=FALSE, include=FALSE}
water=read.csv("data/AGW011.csv")
water[water=="na"] <- NA
npk_water=water[,c(1:9,13,26)]
tmp_water=na.omit(npk_water)
tmp_water[tmp_water=="bdl"] <- 0
table(tmp_water$Location)
water_l1=subset(tmp_water,tmp_water$Location=="4-6Eis1")
water_l2=subset(tmp_water,tmp_water$Location=="4-6Eis2")
water_l3=subset(tmp_water,tmp_water$Location=="3-5Mor")
water_l4=subset(tmp_water,tmp_water$Location=="4-6 Mor")
water1=subset(water_l1,water_l1$Recyear>1995 & water_l1$RECTYPE<1998)
```

::: incremental
1.  **Commutative rule**
    -   Order doesn’t matter
2.  **Associative rule**
    -   Order still doesn’t matter
3.  **Distributive rule**
    -   Order matters sometimes
:::

## Commutative Rule

> When two terms are added$^1$, order does not matter

$$(a + b) = (b + a)$$

> When two terms are multiplied$^2$, order does not matter

$$(a \times b) = (b \times a)$$

::: small
$^1$ subtraction is just negative addition

$^2$ division is just fraction multiplication
:::

## Quick note

**Subtraction is addition**

$$
\begin{aligned}
a - b =  a + (-b) \\
\\
10 - 5 = 5 \\ 
\\
10 + (-5) = 5 \\ 
\\
10 - 5 =  10+(-5) \\
\end{aligned}
$$

## Quick note

**Division is multiplication**

$$
\begin{aligned}
a/b=  a\times 1/b \\
\\
10/5=  2 \\
\\
10 \times 1/5=  2\\
\\
10/5=  10 \times 1/5
\end{aligned}
$$

## Associative Rule

> When $\ge 3$ terms are added, order does not matter

$$a+(b+c)=(a+b)+c $$

> When $\ge 3$ terms are multiplied, order does not matter

$$a \times (b \times c) = (a \times b) \times c$$

## Distributive Rule

> When a term is multiplied over the addition of two terms, the result is an output equal to the sum of their individual products with the first term

$$
\begin{aligned}
a \times (b+c) = (a \times b) + (a \times c) \\ 
\\
5 \times(10+2)=5 \times 12=60\\
\\
(5 \times 10) + (5 \times 2) = 50 + 10 = 60
\end{aligned}
$$

## PEMDAS

::: incremental
1.  Parentheses

2.  Exponents

3.  Multiplication and Division

4.  Addition and Subtraction
:::

## Basic Exponent Properties

$$
\begin{aligned}
& a^m \times a^n= a^{m+n} \\
& \frac{a^m}{a^n}=a^{m-n}\\
& a^0=1\\
& a^{-m}=\frac{1}{a^m}\\
& (a^m)^n=a^{m\times n}\\
& (ab)^m=a^m\times b^m \\
& \left(\frac{a}{b} \right)^m=\frac{a^m}{b^m}
\end{aligned}
$$

## Logarithms

$$
\begin{aligned}
10^3& =1000 \\
\log_{10}(1000)& =3 \\
\\
a^b& =c \\
\log_a(c)& =b \\
\\
\log_a(xy)& =\log_a(x)+\log_a(y)\\
\log_a(x/y)& =\log_a(x)-\log_a(y)\\
\log_a(x^y)& =y\log_a(x)
\end{aligned}
$$

## Equalities

$$
\begin{array}{|c|c|}
\hline
= & \text{Is equal to}\\
\hline
\equiv & \text{Is defined as}\\
\hline
\approx & \text{Is approximately}\\
\hline
< & \text{Is less than}\\
\hline
> & \text{Is greater than}\\
\hline
\le & \text{Is less than or equal to} \\
\hline
\ge & \text{Is greater than or equal to}\\
\hline
\end{array}
$$

## Some Definitions

$$
\begin{aligned}
\pi \approx 3.14 \\
\\
\tau = 2\pi \\
\\
e \approx 2.718 \\
\\
\log = \log_{10}\\
\\
\ln = \log_e \\
\end{aligned}
$$

## Univariate Solutions

$$
\begin{aligned}
15=0.5x+10\\
5=0.5x\\
2.5=x\\
\end{aligned}
$$

$$
\begin{aligned}
22=e^x+2\\
20=e^x\\
\log(20)=x\\
\end{aligned}
$$

# Notation

We're doing Greek letters now

## Denoting Data

$$
\begin{array}{|c|c|}
\hline
n & \text{Sample size} \\
\hline
N & \text{Population size} \\
\hline
x, \ y, \ z & \text{(conventionally) Data} \\
\hline
i,\ j & \text{Indices} \\
\hline
x_i, \ y_j & \text{Indexed data} \\ 
\hline
\text{Lowercase Greek letters} & \text{Population parameters} \\
\hline
\text{Lowercase Arabic letters} & \text{Sample statistics} \\
\hline
\end{array}
$$

## Summation

::: incremental
-   Statisticians are mathematicians

-   Mathematicians like to generalize and simplify

    -   Because we're very lazy

-   Add all of the whole numbers from $1$ to $10$ together

-   $1+2+3+4+5+6+7+8+9+10=55$

-   Inconvenient to write and say
:::

## Summation

::: small
Consider a single "variable", $x_i$, where $x_1=1, x_2=2, ... , x_{10}=10$
:::

$$
\begin{aligned}
& \sum_{i=1}^{10}x_i=1+2+3+4+5+6+7+8+9+10\\
\\
& \sum_{i=1}^{10}x_i=55\\
\end{aligned}
$$

::: incremental
-   Convenient to write and say
:::

## Notes/Warnings

:::: small
::: incremental
-   "In the rabbit hole"

    -   I'll eventually slip up with vocabulary

-   If I say "vector" I mean a line of numbers (column/row in Excel/Google Sheets)

-   If I say "matrix" I mean a box of numbers (a table)

-   If I use any term related to calculus, it's not necessary knowledge, it's an aside

-   If you're confused by a vocab term, raise your hand

    -   $50/50$ I'm just going crazy

    -   $100\%$ chance you aren't alone
:::
::::

# Measures of Center

## Motivation

::: incremental
-   The brick they use to build everything in this town

    -   It's in my water

-   "Ogden Limestone"

-   Kansas doesn't want random stuff in the water

-   Potassium is a good indicator for this limestone

-   How can we determine if the problem is getting worse?
:::

## Motivation

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
```

## Mean

$$\bar x = \frac{1}{n}\sum_{i=1}^nx_i$$

$$\mu = \frac{1}{N}\sum_{i=1}^n x_i$$

::: incremental
-   "Average"

-   Most common measure of center

-   Take all the values in the data, add them up, divide by the number of data points
:::

## Mean

::: small
$$
\begin{array}{|c|c|c|c|c|}
\hline
1.92 & 1.4 & 1.45 & 1.25 & 1.23 \\
\hline
\end{array}
$$

$$
\begin{aligned}
& \bar x = \frac{1}{n}\sum_{i=1}^nx_i\\
\\
& \bar x = \frac{1}{5}(1.92 + 1.4 + 1.45 + 1.25 + 1.23)\\
\\
& \bar x = \frac{1}{5}(7.25)\\
\\
& \bar x = 1.45\\
\end{aligned}
$$
:::

## Resistance

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.92 & 1.4 & 1.45 & 1.25 & 1.23 & \boldsymbol{6.52}\\
\hline
\end{array}
$$

$$\bar x = \frac{1}{6}(7.25+6.52)=2.295$$

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.92 & 1.4 & 1.45 & 1.25 & 1.23 & \boldsymbol{0.01}\\
\hline
\end{array}
$$

$$\bar x = \frac{1}{6}(7.25+0.01)=1.208$$

## Resistance

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(as.numeric(water1$K1)),col="#87755C",lwd=3,lty=2)
text(1.2,6,"Mean",col="#87755C",cex=1.4)
```

## Median

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.92 & 1.4 & 1.45 & 1.25 & 1.23 \\
\hline
\end{array}
$$

::: incremental
-   The simplest measure of center

-   "True" center or middle value

-   Organize data from *lowest* to *highest* value

    -   If $n$ is odd: Choose position ${(n+1)\over2}$ in the ordered data set

    -   If $n$ is even: Pick $n\over 2$ and ${n \over 2}+1$ and average the two data points
:::

## Derivation

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 \\
\hline
\end{array}
$$

$$
\frac{5+1}{2}=\frac{6}{2}=3
$$

## Derivation

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.23 & 1.25 & \boldsymbol{1.4} & 1.45 & 1.92 \\
\hline
\end{array}
$$

$$
\frac{5+1}{2}=\frac{6}{2}=3
$$

$$\text{Median} = 1.4$$

## Derivation

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 & 6.52\\
\hline
\end{array}
$$

$$\frac{6}{2}=3 \quad \& \quad \frac{6}{2}+1=4$$

## Derivation

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.23 & 1.25 & \boldsymbol{1.4} & \boldsymbol{1.45} & 1.92 & 6.52\\
\hline
\end{array}
$$

$$\frac{6}{2}=3 \quad \& \quad \frac{6}{2}+1=4$$

$$\text{Median}=\frac{1.4+1.45}{2}=1.425$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 \\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|}
\hline
\xcancel{1.23} & 1.25 & 1.4 & 1.45 & \xcancel{1.92} \\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & 1.4 & \xcancel{1.45} & \xcancel{1.92} \\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & \boldsymbol{1.4} & \xcancel{1.45} & \xcancel{1.92} \\
\hline
\end{array}
$$

$$\text{Median}=1.4$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 & 6.52\\
\hline
\end{array}
$$

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 & 0.01\\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & 1.25 & 1.4 & 1.45 & 1.92 & \xcancel{6.52}\\
\hline
\end{array}
$$

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & 1.25 & 1.4 & 1.45 & 1.92 & \xcancel{0.01}\\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & 1.4 & 1.45 & \xcancel{1.92} & \xcancel{6.52}\\
\hline
\end{array}
$$

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & 1.4 & 1.45 & \xcancel{1.92} & \xcancel{0.01}\\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & \boldsymbol{1.4} & \boldsymbol{1.45} & \xcancel{1.92} & \xcancel{6.52}\\
\hline
\end{array}
$$

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & \boldsymbol{1.4} & \boldsymbol{1.45} & \xcancel{1.92} & \xcancel{0.01}\\
\hline
\end{array}
$$

$$\text{Median}=\frac{1.4+1.45}{2}=1.425$$

## Skewness

```{r,echo=FALSE,fig.align='center'}
data=c(1,2,2.5,3,3,3.5,3.5,3.5,4,4.5,5,5,5.5,5.5)
par(mar=c(1,0,4,0))
hist(data,breaks=4,xlim=c(0,7),xaxt="n",yaxt="n",
     xlab="n",ylab="n",main="Symmetric: Mean = Median",col="#51288880")

abline(v=median(data),col="#512888",lwd=5,lty=2)
text(2.15,3.5,"Mean",cex=2,col="cyan3")
text(5,3.5,"Median",cex=2,col="red4")
```

## Skewness

```{r,echo=FALSE,fig.align='center'}
data=c(1,1,2,2.5,2.5,2.5,3,3,3.5,3.5,4,5,5,5.5)
par(mar=c(1,0,4,0))
hist(data,breaks=4,xlim=c(0,7),xaxt="n",yaxt="n",
     xlab="n",ylab="n",main="Positively Skewed: Mean > Median",col="#51288880")
abline(v=mean(data),col="cyan3",lwd=5,lty=2)
abline(v=median(data),col="red4",lwd=5,lty=2)
text(4,4,"Mean",cex=2,col="cyan3")
text(1,4,"Median",cex=2,col="red4")
```

## Skewness

```{r,echo=FALSE,fig.align='center'}
data=c(1,2.5,3,3.5,3.5,4,4.5,5,5,5,5,5.5,5.5,5.5)
par(mar=c(1,0,4,0))
hist(data,breaks=4,xlim=c(0,7),xaxt="n",yaxt="n",
     xlab="n",ylab="n",main="Negatively Skewed: Mean < Median",col="#51288880")
abline(v=mean(data),col="cyan3",lwd=5,lty=2)
abline(v=median(data),col="red4",lwd=5,lty=2)
text(3.15,4,"Mean",cex=2,col="cyan3")
text(6,4,"Median",cex=2,col="red4")
```

## Frequency

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(as.numeric(water1$K1)),col="#87755C",lwd=3,lty=2)
abline(v=median(as.numeric(water1$K1)),col="#445B30",lwd=3,lty=2)
legend("topright", 
  legend = c("Mean", "Median"), 
  col = c("#87755C","#445B30"),
  lty=1,lwd=3,
  cex = 1.2, 
  text.col = "black", 
  horiz = F , 
  inset = c(0.1, 0.1))
```

## Frequency

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(as.numeric(water1$K1)),col="#87755C",lwd=3,lty=2)
abline(v=median(as.numeric(water1$K1)),col="#445B30",lwd=3,lty=2)
legend("topright", 
  legend = c("Mean", "Median"), 
  col = c("#87755C","#445B30"),
  lty=1,lwd=3,
  cex = 1.2, 
  text.col = "black", 
  horiz = F , 
  inset = c(0.1, 0.1))
arrows(0.92,12,1.045,8,lwd=3,col="#1E3618")
```

## Mode

$$
\begin{array}{|c|c|c|c|c|}
\hline
\boldsymbol{3} & \boldsymbol{3} & 5 & 7 & 12 \\
\hline
\end{array}
$$

::: incremental
-   Most frequent observation

-   Resistant

-   Not always very informative

-   Easy to skew in a few cases

-   Imagine a homogeneous data set
:::

## Mode

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
set.seed(73)
hist(runif(100000,1,5),xlab="",main="",col="#D4D2C6")
```

# Measures of Spread

## Spread

How can we look at everything going on *between* the data?

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(as.numeric(water1$K1)),col="#87755C",lwd=3,lty=2)
abline(v=median(as.numeric(water1$K1)),col="#445B30",lwd=3,lty=2)
abline(v=1.075,col="#1E3618",lwd=3,lty=2)
legend("topright", 
  legend = c("Mean","Median","Mode"), 
  col = c("#87755C","#445B30","#1E3618"),
  lty=1,lwd=3,
  cex = 1.2, 
  text.col = "black", 
  horiz = F , 
  inset = c(0.1, 0.1))
```

## Range

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
```

## Range

Difference between the largest and smallest data value

$$\text{Range} = \text{Maximum} - \text{Minimum}$$

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
arrows(1.01,8,1.01,4.5,lwd=2,col="red4")
arrows(1.38,8,1.38,2.5,lwd=2,col="red4")
```

## Range

$$\text{Range} = 1.40 - 1.01 = 0.39$$

$$
\begin{array}{|c|c|c|}
\hline
\text{Mean} & \text{Median} & \text{Range} \\
\hline
1.17 & 1.12 & 0.39 \\
\hline
\end{array}
$$

::: incremental
-   Decent context to the data

-   Reduces reliance on graphics
:::

## Range

- Same general information as the graphic

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(as.numeric(water1$K1)),col="#87755C",lwd=3,lty=2)
abline(v=median(as.numeric(water1$K1)),col="#445B30",lwd=3,lty=2)
legend("topright", 
  legend = c("Mean", "Median"), 
  col = c("#87755C","#445B30"),
  lty=1,lwd=3,
  cex = 1.2, 
  text.col = "black", 
  horiz = F , 
  inset = c(0.1, 0.1))
```

## Range

- "How can we break this?"

```{r, echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}
set.seed(1)
dat1=round(rnorm(100,0,1))
dat2=c(rep(1,50),rep(10,50))

par(mar=c(1,1,1,1),mfrow=c(1,2))
hist(dat1,breaks=20,xaxt="n",yaxt="n",xlab="",ylab="",main="",col="#4B9CA290")
hist(dat2,xaxt="n",yaxt="n",xlab="",ylab="",main="",col="#225F0290")
```

## Spread

- How can we describe the degree with which B is *clustered*?

```{r, echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.width="80%",out.height="80%"}
set.seed(10)
A=rnorm(10,3.1,0.8)
B=rnorm(10,3.1,0.3)
par(mfrow=c(1,2))
hist(A,breaks=4,xlim=c(1,5),ylim=c(0,5),col="#c6dc9390",xlab="Aboveground Biomass (g/plant)",main="Clover Species A")
abline(v=mean(A),col="#63a91f",lwd=4,lty=2)
hist(B,breaks=4,xlim=c(1,5),col="#c6dc9390",xlab="Aboveground Biomass (g/plant)",main="Clover Species B")
abline(v=mean(B),col="#63a91f",lwd=4,lty=2)
```

## Variance

- Start with the mean

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(as.numeric(water1$K1)),col="#87755C",lwd=3,lty=2)
```

## Variance

- "Center" the data around the mean

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
k_new=as.numeric(water1$K1)-mean(as.numeric(water1$K1))
hist(k_new,col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(k_new),col="#87755C",lwd=3,lty=2)
```

## Variance

- What happens if we add all of these values?

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
k_new=as.numeric(water1$K1)-mean(as.numeric(water1$K1))
hist(k_new,col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(k_new),col="#87755C",lwd=3,lty=2)
```


## Variance

- Absolute values are nice and intuitive

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
k_new=as.numeric(water1$K1)-mean(as.numeric(water1$K1))
hist(abs(k_new),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(abs(k_new)),col="#87755C",lwd=3,lty=2)
```

## Absolute Values

$$
|x|=\begin{cases}
x, & x\ge 0 \\
-x, & x \lt 0
\end{cases}
$$

- Absolute values are nasty in practice

- Nastier with matrix algebra/calculus

## Variance

- Squaring is easier to do

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
k_new=as.numeric(water1$K1)-mean(as.numeric(water1$K1))
hist(k_new^2,col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(k_new^2),col="#87755C",lwd=3,lty=2)
```

## Variance

- Population

$$\sigma^2=\frac{1}{N}\sum_{i=1}^n(x_i-\mu)^2$$

- Sample

$$s^2=\frac{1}{n-1}\sum_{i=1}^n(x_i-\bar x)^2$$

## Variance

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.92 & 1.4 & 1.45 & 1.25 & 1.23 \\
\hline
\end{array}
$$

$$
\begin{aligned}
\bar x = & \ 1.45\\
s^2=& \ \frac{1}{5}\left((1.92-1.45)^2+(1.4-1.45)^2+(1.45-1.45)^2+(1.25-1.45)^2+(1.23-1.45)^2\right)\\
s^2 = & \ \frac{1}{5}\left((0.47)^2+(-0.05)^2+(0)^2+(-0.2)^2+(0.22)^2 \right)\\
s^2= & \ \frac{1}{5} \left( 0.2209+0.0025 + 0 + 0.04 + 0.0484\right)\\
s^2 = & \ \frac{1}{5}(0.3318) = 0.06236
\end{aligned}
$$

## Variance

- There's a problem here:

$$
\begin{aligned}
\bar x = & \ 1.45 \text{ mg/L}\\
s^2=& \ \frac{1}{5}\left((1.92 \text{ mg/L}-1.45 \text{ mg/L})^2+(1.4 \text{ mg/L}-1.45 \text{ mg/L})^2\\

+ (1.45\text{ mg/L}-1.45\text{ mg/L})^2+(1.25\text{ mg/L}-1.45\text{ mg/L})^2\\
+(1.23\text{ mg/L}-1.45\text{ mg/L})^2\right)\\
s^2 = & \ \frac{1}{5}\left((0.47\text{ mg/L})^2+(-0.05\text{ mg/L})^2+(0\text{ mg/L})^2+(-0.2\text{ mg/L})^2+(0.22\text{ mg/L})^2 \right)\\
s^2= & \ \frac{1}{5} \left( 0.2209 \text{ mg}^2/\text{L}^2+0.0025 \text{ mg}^2/\text{L}^2 + 0 \text{ mg}^2/\text{L}^2 + 0.04 \text{ mg}^2/\text{L}^2 + 0.0484 \text{ mg}^2/\text{L}^2\right)\\
s^2 = & \ \frac{1}{5}(0.3318 \text{ mg}^2/\text{L}^2) = 0.06236 \text{ mg}^2/\text{L}^2
\end{aligned}
$$

## Standard Deviation

$$\sigma = \sqrt{\sigma^2}$$

$$s = \sqrt{s^2}$$

$$s=\sqrt{0.06236 \text{ mg}^2/\text{L}^2} = 0.2497198 \text{ mg/L}$$

- Fundamentally achieves the same thing as absolute difference

- Analytically simpler
