---
title: "Statistics as a Language"
subtitle: "STAT 240 - Fall 2025"
author: "Robert Sholl"
footer:  "[github.io/stat240_f25](https://rmshksu.github.io/stat240_f25/)"
logo: "images/logo.jpg"
# embed-resources: true
format: 
  revealjs: 
    theme: slides.scss
    transition: fade
    slide-number: true
editor: visual
execute:
  freeze: auto
---

# Statistics as a Science

```{r message=FALSE, warning=FALSE, include=FALSE}
tsunami=read.table('data/tsunami.tsv',T,'\t')
tsunami=tsunami[-1,]
tsunami=tsunami[,-1]
raw_tsunami=tsunami
tsunami=data.frame(year=tsunami$Year,
                   month=tsunami$Mo,
                   country=tsunami$Country,
                   lat=tsunami$Latitude,
                   long=tsunami$Longitude,
                   valid=tsunami$Tsunami.Event.Validity,
                   cause=tsunami$Tsunami.Cause.Code,
                   magnitude=tsunami$Earthquake.Magnitude,
                   wave_height=tsunami$Maximum.Water.Height..m.)
og_tsunami=na.omit(tsunami)
tsunami=subset(tsunami,tsunami$valid>2 & tsunami$cause==1)
tsunami=na.omit(tsunami)
extreme=subset(tsunami,tsunami$magnitude>8)
```

## The Scientific Method

::: incremental
-   Where does statistics come into the picture?
:::

$$
\begin{array}{|c|}
\hline
\text{Observe}\\
\hline
\text{Question}\\
\hline
\text{Hypothesize}\\
\hline
\text{Experiment}\\
\hline
\text{Analyze}\\
\hline
\text{Conclude}\\
\hline
\end{array}
$$

## Observe and Question

::: incremental
-   Tsunamis can occur after earthquakes

-   They tend to happen after high magnitude earthquakes

-   *How high magnitude does an earthquake have to be to cause a tsunami?*

-   *How high are the tsunami waves from high magnitude earthquakes?*
:::

## Hypothesize

::: small
> "On 30 July 2025, at 11:24:52 PETT (23:24:52 UTC, 29 July), a $M_w$ 8.8 megathrust earthquake struck off the eastern coast of the Kamchatka Peninsula in the Russian Far East, 119 km (74 mi) east-southeast of the coastal city of Petropavlovsk-Kamchatsky." -- [Wikipedia](https://en.wikipedia.org/wiki/2025_Kamchatka_Peninsula_earthquake)
:::

::: incremental
-   This earthquake was well above the severity to cause a tsunami

-   Higher magnitude earthquakes result in tsunamis with high waves
:::

## Experiment

::: incremental
-   We can't experiment well with this kind of science

-   But we can run a sort of "psuedo"-experiment

-   We select $100$ tsunamis at random and see if they were the product of an earthquake

-   We find out that $68$ of them had some connection to a earthquake
:::

## Vocabulary (Get used to it)

::: incremental
-   **Population:** the *entire collection* of individuals about which information is sought.

-   **Sample:** a *subset* of population, containing the individuals that are actually observed.

-   **Statistics:** is the study of procedures for collecting, describing, and drawing conclusions from information.

    -   the science of describing or making inferences about a population, from a sample.
:::

## Analyze

> Of the 748 tsunamis that have occurred since 1970, 464 of them can be directly associated with an earthquake.

::: incremental
-   From this data we can say two things

-   $62\%$ of tsunamis are caused by earthquakes

-   $68\%$ of the tsunamis in our sample were caused by earthquakes

    -   **Parameter:** a value that describes an entire population.

    -   **Statistic:** a value that describes a sample.
:::

# Sampling Techniques

I hope you're hungry

## Simple Random Sampling

> How clean are the campus dining halls? Let's find out!

::: incremental
-   We decide to line up all of the dining hall workers at Kramer

-   Everyone is assigned a number, $1$ to $50$

-   We randomly select $10$ numbers

-   Then swab the hands of every person who's number came up
:::

## Simple Random Sampling

::: incremental
-   **Simple Random Sample (SRS):** a sample chosen by a method in which collection of n population items is equally likely to make up the sample.

-   Every worker had the same chance of being selected

    -   $1/50$

    -   Note: this only works if we select all $10$ numbers simultaneously

    -   We'll play with that concept more later on
:::

## Stratified Sampling

> Maybe the $10$ workers we selected all work the same station together. How can we make sure we get a sample from every station?

::: incremental
-   Divide up workers by station, $5$ in total

-   Assign each group of workers numbers $1$ through $10$

-   Randomly select $2$ numbers

-   Swab the hands of each of those numbers from each group
:::

## Statified Sampling

::: incremental
-   **Stratified Sample:** The population is divided into groups, called **strata**, where the members of each *stratum* are similar in some way. Then a SRS is drawn from each stratum.

-   We *stratified* by station

-   We defined a SRS for the *strata*

-   We sampled from the strata based off of that SRS
:::

## Cluster Sampling

> Foodborne outbreaks can also happen as a result of contaminated regions, not just bad worker hygeine.

::: incremental
-   We split the dining center up into a grid of 5 ft by 5 ft squares

-   We select $10$ squares at random and swab the entire square

-   Each of these squares are called *clusters*

-   **Cluster Sampling:** Items are drawn from the population in groups, or clusters.
:::

## Systematic Sampling

> At the end of the day, the delivery mechanism of foodborne illness is the food itself.

::: incremental
-   We decide to sample the food coming off the line directly

-   It'd be a massive interruption of service to sample every food item coming out

    -   Plus, we can't serve food once we've sampled it

-   So we decide to sample every $10^{th}$ item that comes from any station for possible contaminants
:::

## Systematic Sampling

<br>

::: incremental
-   Define a number, denoted $k$

-   Sample the $k^{th}$ item that's observed

-   Repeat the sampling process every $k$ items that are observed
:::

## Voluntary Response

::: incremental
-   We design a simple survey that asks the students:

    -   On a scale of 1 to 10 how clean does Kramer dining center appear?

    -   Have you ever felt sick after eating at the dining center?

    -   Do you think the dining center prioritizes food safety adequately?

-   We send the survey to every student that's eaten at Kramer in the past year
:::

## Samples of Convenience

::: incremental
-   How does the class feel about Kramer on a scale of 1 to 10?

-   It an easy sample to acquire, we just used who we had

-   It's a terrible sampling technique in general

-   Why?
:::

# Data Types

## Data

> What is data?

::: incremental
-   **Data:** Information that has been collected

-   **Individual:** *Something* the information has been collected on (People/Places/Things/etc.)

-   **Variables:** Characteristics about the **individuals** we collected information (**data**) from
:::

## Variables

::: {style="font-size: 0.8em;"}
$$
\scriptsize
\begin{array}{|c|c|c|c|c|c|}
\hline
\textbf{Location of harvest} & \textbf{Date of harvest} & \textbf{Sex} & \textbf{Age class} & \textbf{Body mass in kg} \\
\hline
\text{Boyer}  & 2005\text{-}10\text{-}15 & \text{Female} & 3.5 & 34.0 \\
\text{Desoto} & 2004\text{-}12\text{-}12 & \text{Male}   & 3.5 & 71.2 \\
\text{Desoto} & 2009\text{-}10\text{-}17 & \text{Male}   & 0.5 & 21.8 \\
\text{Desoto} & 2010\text{-}01\text{-}02 & \text{Male}   & 0.5 & 19.5 \\
\text{Desoto} & 2005\text{-}12\text{-}11 & \text{Female} & 3.5 & 45.4 \\
\hline
\end{array}
$$
:::

::: incremental
-   Not all variables are held equal

-   What's the difference between column 1 and column 2?

    -   What about 3 and 4?
:::

## Categorical Variables

**Qualitative (Categorical) variable:** The value of the variable represents a *descriptive categories*

::: incremental
-   Identifying labels or names

-   We can't really do math with a label or name

-   "Male" $= 0$, "Female" $= 1$
:::

## Categorical Variables

::: incremental
-   **Ordinal variables:** Categories/values of the variable have a natural ordering

    -   Letter grade: A, B, C, D, F

    -   Age classifications: Young, Middle-aged, Old

-   **Nominal variable:** Categories/values of the variable cannot be ordered **naturally**

    -   Degree program

    -   Sex: Male / Female
:::

## Numeric Variables

**Quantitative variable:** The value of the variable represents a *meaningful number*

::: incremental
-   Body mass, age, height, time

-   We can inherently do math with these

-   Can still be problematic due to *units*
:::

## Numeric Variables

::: incremental
-   **Discrete variable:** A countable number of values (0, 1, 2, 3, 4, ...)

    -   Population size of fish in a pond

    -   How many times a coin flip was successfully called

-   **Continuous variable:** A continuous range of numbers (0, 0.1, 0.11, 0.111, ...)

    -   Temperature

    -   Volume of liquid in a glass

    -   Body Mass
:::

## Numeric Variables

::: incremental
-   **Interval** level

    -   Differences between values make sense

    -   Ratios don't make sense because **zero has no meaning**

-   **Ratio** level

    -   Differences between values make sense

    -   Ratios **also make sense**

    -   Zero **has meaning**, it represents absence of the quantity
:::

# Summarizing Data

## Frequency Distribution

> Statistics is really good at summarizing grotesque amounts of information

::: incremental
-   Of the $748$ tsunamis that have occurred since 1970, when do they usually occur?

-   We can look at the months, since exact date is a little impractical

-   Instead of a table of raw data, we can just record each time a specific month appears
:::

## Frequency Distribution

**Frequency distribution:**

-   Groups data into categories

-   Records the number of observations that fall into each category

-   "How **frequently** do these variables occur in my sample?"

::: {style="font-size: 0.8em;"}
$$
\begin{array}{|c|c|c|c|c|c|c|}
\hline
\text{Month} & \text{Jan} & \text{Feb} & \text{Mar} & \text{Apr} & \text{May} & \text{Jun} \\
\hline
\text{Frequency} & 64 & 53 & 68 & 62 & 62 & 63\\
\hline
\text{Month} & \text{Jul} & \text{Aug} & \text{Sep} & \text{Oct} & \text{Nov} & \text{Dec}\\
\hline
\text{Frequency} & 66 & 67 & 58 & 63 & 65 & 57 \\
\hline
\end{array}
$$
:::

## Relative Frequency

::: small
**Relative frequency distribution**

-   Divide the frequency by the total observations

-   This gives us the **proportion** of units in each category

-   "What **percentage** of my sample is this category?"
:::

::: {style="font-size: 0.8em;"}
$$
\begin{array}{|c|c|c|c|c|c|c|}
\hline
\text{Month} & \text{Jan} & \text{Feb} & \text{Mar} & \text{Apr} & \text{May} & \text{Jun} \\
\hline
\text{Frequency} & 64 & 53 & 68 & 62 & 62 & 63\\
\hline
\text{Relative Frequency} & 0.09 & 0.07 & 0.09 & 0.08 & 0.08 & 0.08\\
\hline
\text{Month} & \text{Jul} & \text{Aug} & \text{Sep} & \text{Oct} & \text{Nov} & \text{Dec}\\
\hline
\text{Frequency} & 66 & 67 & 58 & 63 & 65 & 57 \\
\hline
\text{Relative Frequency} & 0.09 & 0.09 & 0.08 & 0.08 & 0.09 & 0.08 \\ 
\hline
\end{array}
$$
:::

## Numeric Frequency

::: incremental
-   We can do the same thing with numeric values

-   If we wanted to group the tsunamis caused by earthquakes by magnitude of earthquake

-   We'd need to define some kind of classification for earthquake magnitude

-   Then we could sort them by that

-   Let's say: $0.0-0.9$, $1.0-1.9$, ...
:::

## Numeric Frequency

**There were no tsunamis caused by earthquakes less than a magnitude of 4, so we can remove those classes**

<br>

::: {style="font-size: 0.6em;"}
$$
\begin{array}{|c|c|c|c|c|c|c|}
\hline
\text{Class} &  4.0-4.9 & 5.0-5.9 & 6.0-6.9 & 7.0-7.9 & 8.0-8.9 & 9.0-9.9 \\
\hline
\text{Frequency} &  2  & 22 & 188 & 225 & 25 & 2\\
\hline
\text{Relative Frequency} & 0.0043 & 0.0474 & 0.4052 & 0.4849 & 0.0539 & 0.0043\\
\hline
\end{array}
$$
:::

::: incremental
-   Same table, same general process, same interpretation

-   Class definitions are arbitrary but try to make them *make sense*
:::

# Visualizing Data

## Bar Graphs

```{r,echo=FALSE,fig.align='center'}
Mo_tsunami=as.data.frame(table(raw_tsunami$Mo))
Mo_tsunami$Var1=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
par(mar = c(3, 4, 3, 1))
barplot(Mo_tsunami$Freq,
        names.arg = Mo_tsunami$Var1,
        las = 1,
        cex.names = 0.8,
        col = "#00CBD090",
        ylim=c(0,70),
        main = "Occurrence of Tsunamis by Month",
        xlab = "Month")
```

## Bar Graphs

```{r,echo=FALSE,fig.align='center'}
par(mar = c(5, 5, 4, 2))
barplot(Mo_tsunami$Freq,
        names.arg = Mo_tsunami$Var1,
        horiz = T,
        las = 1,
        cex.names = 0.8,
        col = "#00CBD090",
        xlim=c(0,70),
        main = "Occurrence of Tsunamis by Month",
        xlab = "Month")
```

## Pie Chart

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
library(scales)
library(ggplot2)
tab_lev=Mo_tsunami
tab_lev$Relative_Frequency=round(tab_lev$Freq/sum(tab_lev$Freq),digits=2)
ggplot(data=tab_lev, aes(x="", y=Freq, fill=Var1)) +
  geom_bar(stat="identity",width=1,color="black") +
  labs(x="Month", y="Frequency", title="Occurrence of Tsunamis by Month", fill="Month") +
  geom_text(aes(label = percent(Relative_Frequency)),
            position = position_stack(vjust = 0.5)) +
  coord_polar("y",start=0) + theme_void()
```

## Histograms

::: {style="font-size: 0.6em;"}
$$
\begin{array}{|c|c|c|c|c|c|c|}
\hline
\text{Class} &  4.0-4.9 & 5.0-5.9 & 6.0-6.9 & 7.0-7.9 & 8.0-8.9 & 9.0-9.9 \\
\hline
\text{Frequency} &  2  & 22 & 188 & 225 & 25 & 2\\
\hline
\text{Relative Frequency} & 0.0043 & 0.0474 & 0.4052 & 0.4849 & 0.0539 & 0.0043\\
\hline
\end{array}
$$
:::

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(tsunami$magnitude,breaks=6,xlim=c(4,10),col="#C1E2CC90",xlab="Magnitude",main="Earthquakes associated with tsunamis")
```

## Histograms

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(tsunami$magnitude,breaks=10,xlim=c(4,10),col="#C1E2CC90",xlab="Magnitude",main="Earthquakes associated with tsunamis")
```

## Histograms

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(tsunami$magnitude,breaks=60,col="#C1E2CC90",xlab="Magnitude",main="Earthquakes associated with tsunamis")
```

## Time plots

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
seven=subset(tsunami,tsunami$magnitude>8)
plot(seven$year,seven$wave_height,type="b",cex=1.2,lwd=2,xlab="Year",ylab="Wave Height",main="Wave Heights of High Magnitude Earthquakes (>8.0)")
```

# Go Away
