---
title: "Summary Statistics"
subtitle: "STAT 240 - Fall 2025"
author: "Robert Sholl"
footer:  "[github.io/stat240_f25](https://rmshksu.github.io/stat240_f25/)"
logo: "images/logo.jpg"
format: 
  revealjs: 
    theme: slides.scss
    transition: fade
    slide-number: true
editor: visual
execute:
  freeze: auto
---

<script type="text/x-mathjax-config">
MathJax.Hub.Register.StartupHook("TeX Jax Ready",function () {
  MathJax.Hub.Insert(MathJax.InputJax.TeX.Definitions.macros,{
    cancel: ["Extension","cancel"],
    bcancel: ["Extension","cancel"],
    xcancel: ["Extension","cancel"],
    cancelto: ["Extension","cancel"]
  });
});
</script>

```{r message=FALSE, warning=FALSE, include=FALSE}
water=read.csv("data/AGW011.csv")
water[water=="na"] <- NA
npk_water=water[,c(1:9,13,26)]
tmp_water=na.omit(npk_water)
tmp_water[tmp_water=="bdl"] <- 0
table(tmp_water$Location)
water_l1=subset(tmp_water,tmp_water$Location=="4-6Eis1")
water_l2=subset(tmp_water,tmp_water$Location=="4-6Eis2")
water_l3=subset(tmp_water,tmp_water$Location=="3-5Mor")
water_l4=subset(tmp_water,tmp_water$Location=="4-6 Mor")
water1=subset(water_l1,water_l1$Recyear>1995 & water_l1$RECTYPE<1998)
```

# Math Review

## Rules

::: incremental

1. **Commutative rule**
   - Order doesn’t matter  

<span style="display:block; height:0.5em;"></span>

2. **Associative rule**
   - Order still doesn’t matter  

<span style="display:block; height:0.5em;"></span>

3. **Distributive rule**
   - Order matters sometimes

:::

## Commutative Rule

>When two terms are added$^1$, order does not matter

$$(a + b) = (b + a)$$

<span style="display:block; height:0.5em;"></span>

>When two terms are multiplied$^2$, order does not matter

$$(a \times b) = (b \times a)$$

<span style="display:block; height:0.5em;"></span>


:::{.small}

$^1$ subtraction is just negative addition

$^2$ division is just fraction multiplication

:::


## Quick note

**Subtraction is addition**

$$
\begin{aligned}
a - b =  a + (-b) \\
\\
10 - 5 = 5 \\ 
\\
10 + (-5) = 5 \\ 
\\
10 - 5 =  10+(-5) \\
\end{aligned}
$$

## Quick note

**Division is multiplication**

$$
\begin{aligned}
a/b=  a\times 1/b \\
\\
10/5=  2 \\
\\
10 \times 1/5=  2\\
\\
10/5=  10 \times 1/5
\end{aligned}
$$


## Associative Rule

>When $\ge 3$ terms are added, order does not matter

$$a+(b+c)=(a+b)+c $$

<span style="display:block; height:0.5em;"></span>

>When $\ge 3$ terms are multiplied, order does not matter

$$a \times (b \times c) = (a \times b) \times c$$

## Distributive Rule

>When a term is multiplied over the addition of two terms, the result is an output equal to the sum of their individual products with the first term

$$
\begin{aligned}
a \times (b+c) = (a \times b) + (a \times c) \\ 
\\
5 \times(10+2)=5 \times 12=60\\
\\
(5 \times 10) + (5 \times 2) = 50 + 10 = 60
\end{aligned}
$$

## PEMDAS

::: incremental

1. Parentheses

<span style="display:block; height:0.5em;"></span>

2. Exponents

<span style="display:block; height:0.5em;"></span>

3. Multiplication and Division

<span style="display:block; height:0.5em;"></span>

4. Addition and Subtraction

:::

## Basic Exponent Properties

$$
\begin{aligned}
& a^m \times a^n= a^{m+n} \\
& \frac{a^m}{a^n}=a^{m-n}\\
& a^0=1\\
& a^{-m}=\frac{1}{a^m}\\
& (a^m)^n=a^{m\times n}\\
& (ab)^m=a^m\times b^m \\
& \left(\frac{a}{b} \right)^m=\frac{a^m}{b^m}
\end{aligned}
$$

## Logarithms

$$
\begin{aligned}
10^3& =1000 \\
\log_{10}(1000)& =3 \\
\\
a^b& =c \\
\log_a(c)& =b \\
\\
\log_a(xy)& =\log_a(x)+\log_a(y)\\
\log_a(x/y)& =\log_a(x)-\log_a(y)\\
\log_a(x^y)& =y\log_a(x)
\end{aligned}
$$

## Equalities

$$
\begin{array}{|c|c|}
\hline
= & \text{Is equal to}\\
\hline
\equiv & \text{Is defined as}\\
\hline
\approx & \text{Is approximately}\\
\hline
< & \text{Is less than}\\
\hline
> & \text{Is greater than}\\
\hline
\le & \text{Is less than or equal to} \\
\hline
\ge & \text{Is greater than or equal to}\\
\hline
\end{array}
$$



## Some Definitions

$$
\begin{aligned}
\pi \approx 3.14 \\
\\
\tau = 2\pi \\
\\
e \approx 2.718 \\
\\
\log = \log_{10}\\
\\
\ln = \log_e \\
\end{aligned}
$$

## Univariate Solutions

$$
\begin{aligned}
15=0.5x+10\\
5=0.5x\\
2.5=x\\
\end{aligned}
$$

<span style="display:block; height:0.5em;"></span>

$$
\begin{aligned}
22=e^x+2\\
20=e^x\\
\log(20)=x\\
\end{aligned}
$$

# Notation

We're doing Greek letters now

## Denoting Data

$$
\begin{array}{|c|c|}
\hline
n & \text{Sample size} \\
\hline
N & \text{Population size} \\
\hline
x, \ y, \ z & \text{(conventionally) Data} \\
\hline
i,\ j & \text{Indices} \\
\hline
x_i, \ y_j & \text{Indexed data} \\ 
\hline
\text{Lowercase Greek letters} & \text{Population parameters} \\
\hline
\text{Lowercase Arabic letters} & \text{Sample statistics} \\
\hline
\end{array}
$$

## Summation

::: incremental

- Statisticians are mathematicians

<span style="display:block; height:0.3em;"></span>

- Mathematicians like to generalize and simplify

    - Because we're very lazy
    
<span style="display:block; height:0.3em;"></span>

- Add all of the whole numbers from $1$ to $10$ together

- $1+2+3+4+5+6+7+8+9+10=55$

<span style="display:block; height:0.3em;"></span>

- Inconvenient to write and say

:::

## Summation

:::{.small}

Consider a single "variable", $x_i$, where $x_1=1, x_2=2, ... , x_{10}=10$

:::

$$
\begin{aligned}
& \sum_{i=1}^{10}x_i=1+2+3+4+5+6+7+8+9+10\\
\\
& \sum_{i=1}^{10}x_i=55\\
\end{aligned}
$$

::: incremental

- Convenient to write and say

:::

## Notes/Warnings

::: {.small}

::: incremental

- "In the rabbit hole"

    - I'll eventually slip up with vocabulary
    
<span style="display:block; height:0.3em;"></span>
    
- If I say "vector" I mean a line of numbers (column/row in Excel/Google Sheets)

- If I say "matrix" I mean a box of numbers (a table)

- If I use any term related to calculus, it's not necessary knowledge, it's an aside

<span style="display:block; height:0.3em;"></span>

- If you're confused by a vocab term, raise your hand

    - $50/50$ I'm just going crazy

    - $100\%$ chance you aren't alone

:::

:::

# Measures of Center

## Motivation

::: incremental

- The brick they use to build everything in this town

    - It's in my water

- "Ogden Limestone"
    
- Kansas doesn't want random stuff in the water

- Potassium is a good indicator for this limestone

- How can we determine if the problem is getting worse?

:::

## Motivation

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
```

## Mean

$$\bar x = \frac{1}{n}\sum_{i=1}^nx_i$$

$$\mu = \frac{1}{N}\sum_{i=1}^n x_i$$

::: incremental

- "Average"

- Most common measure of center

- Take all the values in the data, add them up, divide by the number of data points

:::

## Mean

:::{.small}

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.92 & 1.4 & 1.45 & 1.25 & 1.23 \\
\hline
\end{array}
$$

$$
\begin{aligned}
& \bar x = \frac{1}{n}\sum_{i=1}^nx_i\\
\\
& \bar x = \frac{1}{5}(1.92 + 1.4 + 1.45 + 1.25 + 1.23)\\
\\
& \bar x = \frac{1}{5}(7.25)\\
\\
& \bar x = 1.45\\
\end{aligned}
$$

:::

## Resistance

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.92 & 1.4 & 1.45 & 1.25 & 1.23 & \boldsymbol{6.52}\\
\hline
\end{array}
$$

$$\bar x = \frac{1}{6}(7.25+6.52)=2.295$$

<span style="display:block; height:0.5em;"></span>

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.92 & 1.4 & 1.45 & 1.25 & 1.23 & \boldsymbol{0.01}\\
\hline
\end{array}
$$

$$\bar x = \frac{1}{6}(7.25+0.01)=1.208$$

## Resistance

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(as.numeric(water1$K1)),col="#87755C",lwd=3,lty=2)
text(1.2,6,"Mean",col="#87755C",cex=1.4)
```

## Median

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.92 & 1.4 & 1.45 & 1.25 & 1.23 \\
\hline
\end{array}
$$

::: incremental

- The simplest measure of center

- "True" center or middle value

- Organize data from *lowest* to *highest* value

    - If $n$ is odd: Choose position ${(n+1)\over2}$ in the ordered data set

    - If $n$ is even: Pick $n\over 2$ and ${n \over 2}+1$ and average the two data points

:::

## Derivation

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 \\
\hline
\end{array}
$$



<span style="display:block; height:0.5em;"></span>


$$
\frac{5+1}{2}=\frac{6}{2}=3
$$

## Derivation

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.23 & 1.25 & \boldsymbol{1.4} & 1.45 & 1.92 \\
\hline
\end{array}
$$



<span style="display:block; height:0.5em;"></span>


$$
\frac{5+1}{2}=\frac{6}{2}=3
$$

<span style="display:block; height:0.5em;"></span>

$$\text{Median} = 1.4$$

## Derivation

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 & 6.52\\
\hline
\end{array}
$$


<span style="display:block; height:0.5em;"></span>

$$\frac{6}{2}=3 \quad \& \quad \frac{6}{2}+1=4$$

## Derivation

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.23 & 1.25 & \boldsymbol{1.4} & \boldsymbol{1.45} & 1.92 & 6.52\\
\hline
\end{array}
$$


<span style="display:block; height:0.5em;"></span>

$$\frac{6}{2}=3 \quad \& \quad \frac{6}{2}+1=4$$

<span style="display:block; height:0.5em;"></span>

$$\text{Median}=\frac{1.4+1.45}{2}=1.425$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 \\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|}
\hline
\xcancel{1.23} & 1.25 & 1.4 & 1.45 & \xcancel{1.92} \\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & 1.4 & \xcancel{1.45} & \xcancel{1.92} \\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & \boldsymbol{1.4} & \xcancel{1.45} & \xcancel{1.92} \\
\hline
\end{array}
$$

$$\text{Median}=1.4$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 & 6.52\\
\hline
\end{array}
$$

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
1.23 & 1.25 & 1.4 & 1.45 & 1.92 & 0.01\\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & 1.25 & 1.4 & 1.45 & 1.92 & \xcancel{6.52}\\
\hline
\end{array}
$$

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & 1.25 & 1.4 & 1.45 & 1.92 & \xcancel{0.01}\\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & 1.4 & 1.45 & \xcancel{1.92} & \xcancel{6.52}\\
\hline
\end{array}
$$

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & 1.4 & 1.45 & \xcancel{1.92} & \xcancel{0.01}\\
\hline
\end{array}
$$

## Elimination

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & \boldsymbol{1.4} & \boldsymbol{1.45} & \xcancel{1.92} & \xcancel{6.52}\\
\hline
\end{array}
$$

$$
\begin{array}{|c|c|c|c|c|c|}
\hline
\xcancel{1.23} & \xcancel{1.25} & \boldsymbol{1.4} & \boldsymbol{1.45} & \xcancel{1.92} & \xcancel{0.01}\\
\hline
\end{array}
$$

<span style="display:block; height:0.5em;"></span>

$$\text{Median}=\frac{1.4+1.45}{2}=1.425$$

## Skewness

```{r,echo=FALSE,fig.align='center'}
data=c(1,2,2.5,3,3,3.5,3.5,3.5,4,4.5,5,5,5.5,5.5)
par(mar=c(1,0,4,0))
hist(data,breaks=4,xlim=c(0,7),xaxt="n",yaxt="n",
     xlab="n",ylab="n",main="Symmetric: Mean = Median",col="#51288880")

abline(v=median(data),col="#512888",lwd=5,lty=2)
text(2.15,3.5,"Mean",cex=2,col="cyan3")
text(5,3.5,"Median",cex=2,col="red4")
```

## Skewness

```{r,echo=FALSE,fig.align='center'}
data=c(1,1,2,2.5,2.5,2.5,3,3,3.5,3.5,4,5,5,5.5)
par(mar=c(1,0,4,0))
hist(data,breaks=4,xlim=c(0,7),xaxt="n",yaxt="n",
     xlab="n",ylab="n",main="Positively Skewed: Mean > Median",col="#51288880")
abline(v=mean(data),col="cyan3",lwd=5,lty=2)
abline(v=median(data),col="red4",lwd=5,lty=2)
text(4,4,"Mean",cex=2,col="cyan3")
text(1,4,"Median",cex=2,col="red4")
```

## Skewness

```{r,echo=FALSE,fig.align='center'}
data=c(1,2.5,3,3.5,3.5,4,4.5,5,5,5,5,5.5,5.5,5.5)
par(mar=c(1,0,4,0))
hist(data,breaks=4,xlim=c(0,7),xaxt="n",yaxt="n",
     xlab="n",ylab="n",main="Negatively Skewed: Mean < Median",col="#51288880")
abline(v=mean(data),col="cyan3",lwd=5,lty=2)
abline(v=median(data),col="red4",lwd=5,lty=2)
text(3.15,4,"Mean",cex=2,col="cyan3")
text(6,4,"Median",cex=2,col="red4")
```

## Frequency

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(as.numeric(water1$K1)),col="#87755C",lwd=3,lty=2)
abline(v=median(as.numeric(water1$K1)),col="#445B30",lwd=3,lty=2)
legend("topright", 
  legend = c("Mean", "Median"), 
  col = c("#87755C","#445B30"),
  lty=1,lwd=3,
  cex = 1.2, 
  text.col = "black", 
  horiz = F , 
  inset = c(0.1, 0.1))
```

## Frequency

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
hist(as.numeric(water1$K1),col="#92BBBD90",xlab="Potassium (mg/L)",main="Watershed Potassium Levels (1996-1997)")
abline(v=mean(as.numeric(water1$K1)),col="#87755C",lwd=3,lty=2)
abline(v=median(as.numeric(water1$K1)),col="#445B30",lwd=3,lty=2)
legend("topright", 
  legend = c("Mean", "Median"), 
  col = c("#87755C","#445B30"),
  lty=1,lwd=3,
  cex = 1.2, 
  text.col = "black", 
  horiz = F , 
  inset = c(0.1, 0.1))
arrows(0.92,12,1.045,8,lwd=3,col="#1E3618")
```

## Mode

$$
\begin{array}{|c|c|c|c|c|}
\hline
\boldsymbol{3} & \boldsymbol{3} & 5 & 7 & 12 \\
\hline
\end{array}
$$

::: incremental

- Most frequent observation

- Resistant

- Not always very informative

- Easy to skew in a few cases

- Imagine a homogeneous data set

:::

## Mode

```{r,echo=FALSE,warning=FALSE,message=FALSE, fig.align='center'}
set.seed(73)
hist(runif(100000,1,5),xlab="",main="",col="#D4D2C6")
```